---
title: "Substance Abuse"
author: "Andre Williams"
date: '2022-07-29'
output:
  html_document: default
---

# Predicators of Drug Overdose Deaths

# Introduction

Where do most drug overdose deaths occur and what are some characteristics of those areas? Some of the socio-economic factors that were looked into were the median household income, high school completion rate, some college experience rate, and unemployment rate per state. Also some information about the most used drug in a drug overdose death.

# Data

__County Health Rankings: State, High School Completion, Some College Experience, Unemployment__

```{r include=FALSE, echo=FALSE}
library(tidyverse)
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
county_rankings <- read.csv("~/SURE-2022/county_rankings.csv")
income <- read.csv("~/SURE-2022/MedianHouseholdIncome.csv")
```

```{r include=FALSE, echo=FALSE}
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))
colnames(d3)[1] <- 'state'

OD <- d3 %>%
  arrange(state) %>%
  filter(Year %in% "2019",
         Indicator %in% "Number of Drug Overdose Deaths",
         !state %in% c("DC", "YC", "US")) %>%
  select(state, Indicator, Data.Value) %>%
  group_by(state) %>%
  summarise(OD_deaths = sum(Data.Value))
```

```{r echo=FALSE}
county_rankings <- county_rankings %>%
  arrange(State.Abbreviation) %>%
  filter(County.FIPS.Code %in% 0,
         !State.Abbreviation %in% c("US", "DC")) %>%
  select(State.Abbreviation, Unemployment.raw.value, Some.college.raw.value, High.school.completion.raw.value) %>%
  mutate(Unemployment.raw.value = (Unemployment.raw.value * 100),
         High.school.completion.raw.value = (High.school.completion.raw.value * 100),
         Some.college.raw.value = (Some.college.raw.value * 100))

colnames(county_rankings)[1] <- 'state'
```

```{r echo=FALSE}
colnames(income)[1] <- 'state'
income <- income %>%
  mutate(state = fct_recode(state,
                            "MD" = "Maryland",
                            "NJ" = "New Jersey",
                            "HI" = "Hawaii",
                            "MA" = "Massachusetts",
                            "CT" = "Connecticut",
                            "AK" = "Alaska",
                            "NH" = "New Hampshire",
                            "CA" = "California",
                            "VA" = "Virginia",
                            "WA" = "Washington",
                            "CO" = "Colorado",
                            "UT" = "Utah",
                            "MN" = "Minnesota",
                            "NY" = "New York",
                            "DE" = "Delaware",
                            "RI" = "Rhode Island",
                            "IL" = "Illinois",
                            "ND" = "North Dakota",
                            "WY" = "Wyoming",
                            "OR" = "Oregon",
                            "VT" = "Vermont",
                            "TX" = "Texas",
                            "WI" = "Wisconsin",
                            "PA" = "Pennsylvania",
                            "NE" = "Nebraska",
                            "IA" = "Iowa",
                            "NV" = "Nevada",
                            "KS" = "Kansas",
                            "AZ" = "Arizona",
                            "GA" = "Georgia",
                            "SD" = "South Dakota",
                            "ME" = "Maine",
                            "MI" = "Michigan",
                            "OH" = "Ohio",
                            "IN" = "Indiana",
                            "ID" = "Idaho",
                            "FL" = "Florida",
                            "MO" = "Missouri",
                            "MT" = "Montana",
                            "NC" = "North Carolina",
                            "TN" = "Tennessee",
                            "SC" = "South Carolina",
                            "OK" = "Oklahoma",
                            "KY" = "Kentucky",
                            "AL" = "Alabama",
                            "NM" = "New Mexico",
                            "LA" = "Louisiana",
                            "AR" = "Arkansas",
                            "WV" = "West Virginia",
                            "MS" = "Mississippi"))
```

__Median Household Income: State, Household Income__

```{r echo=FALSE}
library(usmap)
library(ggplot2)

plot_usmap(data = income, values = "HouseholdIncome", color = "red") + 
  scale_fill_continuous(low = "white", high = "red", name = "Income", label = scales::comma) +
  labs(title = "Median Household Income by State") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))

```

```{r echo=FALSE, include=FALSE}
library(dplyr)

first <- dplyr::left_join(x = OD, y = county_rankings, by = 'state')
combined <- dplyr::left_join(x = first, y = income, by = 'state')
```

__12 Month-ending Provisional Number and Percent Change of Drug Overdose Deaths: State, Indicator, Data Value__

```{r echo=FALSE}
plot_usmap(data = combined, values = "OD_deaths", color = "red") + 
  scale_fill_continuous(low = "yellow", high = "darkred", name = "Drug OD Deaths", label = scales::comma) +
  labs(title = "Total Drug Overdose Deaths per State") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))
```

```{r echo=FALSE, warning = FALSE, message=FALSE}
library(ggrepel)
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))
d3[is.na(d3)] = 0

d3 <- d3 %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                        "Number of Drug Overdose Deaths"),
         State %in% "CA") %>%
  select(State, Indicator, Data.Value) %>%
  group_by(State, Indicator) %>%
  summarise(Data.Value = sum(Data.Value)) %>%
  arrange(desc(Data.Value))

df2 <- d3 %>% 
  mutate(csum = rev(cumsum(rev(Data.Value))), 
         pos = Data.Value/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Data.Value/2, pos))

d3 %>%
  ggplot(aes(x = "" , y = Data.Value, 
             fill = fct_inorder(str_wrap(Indicator, 20)))) +
  labs(title = "California's Leading Drug in Overdose Deaths") +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Paired") +
  geom_label_repel(data = df2,
                   aes(y = pos, label = paste0(Data.Value)),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Drug Type", ncol = 2, 
                             title.position = "top",
                             title.theme = element_text(size = 10),
                             label.theme = element_text(size = 7))) +
  theme_void() + 
  theme(legend.position = "right")
```

```{r echo=FALSE, warning = FALSE, message=FALSE}
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))
d3[is.na(d3)] = 0

d3 <- d3 %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                        "Number of Drug Overdose Deaths"),
         State %in% "FL") %>%
  select(State, Indicator, Data.Value) %>%
  group_by(State, Indicator) %>%
  summarise(Data.Value = sum(Data.Value)) %>%
  arrange(desc(Data.Value))

df2 <- d3 %>% 
  mutate(csum = rev(cumsum(rev(Data.Value))), 
         pos = Data.Value/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Data.Value/2, pos))

d3 %>%
  ggplot(aes(x = "" , y = Data.Value, 
             fill = fct_inorder(str_wrap(Indicator, 20)))) +
  labs(title = "Florida's Leading Drug in Overdose Deaths") +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Paired") +
  geom_label_repel(data = df2,
                   aes(y = pos, label = paste0(Data.Value)),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Indicator", ncol = 2, 
                             title.position = "top",
                             title.theme = element_text(size = 10),
                             label.theme = element_text(size = 7))) +
  theme_void() + 
  theme(legend.position = "right")
```

```{r echo=FALSE, warning = FALSE, message=FALSE}
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))
d3[is.na(d3)] = 0

d3 <- d3 %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                        "Number of Drug Overdose Deaths"),
         State %in% "OH") %>%
  select(State, Indicator, Data.Value) %>%
  group_by(State, Indicator) %>%
  summarise(Data.Value = sum(Data.Value)) %>%
  arrange(desc(Data.Value))

df2 <- d3 %>% 
  mutate(csum = rev(cumsum(rev(Data.Value))), 
         pos = Data.Value/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Data.Value/2, pos))

d3 %>%
  ggplot(aes(x = "" , y = Data.Value, 
             fill = fct_inorder(str_wrap(Indicator, 20)))) +
  labs(title = "Ohio's Leading Drug in Overdose Deaths") +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Paired") +
  geom_label_repel(data = df2,
                   aes(y = pos, label = paste0(Data.Value)),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Indicator", ncol = 2, 
                             title.position = "top",
                             title.theme = element_text(size = 10),
                             label.theme = element_text(size = 7))) +
  theme_void() + 
  theme(legend.position = "right")
```

# Methods

__K-Means Clustering__

We implemented K-means clustering to determine if there were any correlation between a drug overdose death and the socio-economic factors listed. The reasons for using these four factors was because of their pre-existing relations to someone who uses drugs.
People who are unemployed aren't necessarily not working because they don't want. It can be very stressful looking for a job, and when people are stressed, they look for ways to relieve it and drugs is a common solution.
In most communities, people who did not finish high school tend go down the route of drugs and potentially become an addict.
Drugs of all kinds are present on college campuses and dents tend to indulge in taking drugs because there is immense peer pressure and no supervision.
Household incomes and it's relationship towards drug abuse can vary, especially based on the area.

# Results

Since we grouped our data by states, we will look at which states share similarities and which of the factors do these states fall under.

```{r echo=FALSE}
library(ggrepel)
combined <- combined %>%
  mutate(std_death = as.numeric(scale(OD_deaths, center = TRUE, scale = TRUE)),
         std_unemploy = as.numeric(scale(Unemployment.raw.value, center = TRUE, scale = TRUE)),
         std_hs = as.numeric(scale(High.school.completion.raw.value, center = TRUE, scale = TRUE)),
         std_college = as.numeric(scale(Some.college.raw.value, center = TRUE, scale = TRUE)),
         std_income = as.numeric(scale(HouseholdIncome, center = TRUE, scale = TRUE)))
```

```{r echo=FALSE,warning=FALSE, message=FALSE}
std_kmeans <- 
  kmeans(dplyr::select(combined, std_death, std_unemploy),
         algorithm = "Lloyd", centers = 4, nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(std_kmeans$cluster)) %>%
  ggplot(aes(x = std_death, y = std_unemploy,
             color = state_clusters, label = state)) +
  labs(x = "Total Deaths", y = "Unemployment Rate",
       title = "Cluster of Deaths and Unemployment Rate by State") +
  geom_point() +
  geom_text_repel(size = 2) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5)) +
  coord_fixed()

```

Here we can see a common factor in where states share similarities among unemployment rate. However, there is some correlation between unemployment rates and states with a lower drug overdose death. States with a lower unemployment rate tend to have a lower number of total deaths as seen as in the green and black clusters.

```{r echo=FALSE,warning=FALSE, message=FALSE}
std_kmeans <- 
  kmeans(dplyr::select(combined, std_death, std_hs),
         algorithm = "Lloyd", centers = 4, nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(std_kmeans$cluster)) %>%
  ggplot(aes(x = std_death, y = std_hs,
             color = state_clusters, label = state)) +
  labs(x = "Total Deaths", y = "HS Completion",
       title = "Cluster of Deaths and HS Completion Rate by State") +
  geom_point() + 
  geom_text_repel(size = 2) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5)) +
  coord_fixed()

```

There is wide spectrum of states that fall within the high school completion rate range. States with a higher high school completion rate tend to have a lower number of drug overdose deaths. But the data becomes tricky when examining the green and yellow cluster of states because they have a more common amount of total deaths, yet has a large range of high school completion rate. And most states with a high number of deaths lean towards the median of the HS completion rate.

```{r echo=FALSE,warning=FALSE, message=FALSE}
std_kmeans <- 
  kmeans(dplyr::select(combined, std_death, std_college),
         algorithm = "Lloyd", centers = 4, nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(std_kmeans$cluster)) %>%
  ggplot(aes(x = std_death, y = std_college,
             color = state_clusters, label = state)) +
  labs(x = "Total Deaths", y = "College Experience",
       title = "Cluster of Deaths and College Experience Rate by State") +
  geom_point() + 
  geom_text_repel(size = 2) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5)) +
  coord_fixed()

```

Unlike the other two clusters, having some college experience seems to not have a clear or slightly visible effect on a drug overdose death.

```{r echo=FALSE,warning=FALSE, message=FALSE}
std_kmeans <- 
  kmeans(dplyr::select(combined, std_death, std_income),
         algorithm = "Lloyd", centers = 4, nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(std_kmeans$cluster)) %>%
  ggplot(aes(x = std_death, y = std_income,
             color = state_clusters, label = state)) +
  labs(x = "Total Deaths", y = "Household Income",
       title = "Cluster of Deaths and Median Household Income by State") +
  geom_point() + 
  geom_text_repel(size = 2) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5)) +
  coord_fixed()

```

Household income plays a little bit of a factor here as states with lower median household income tend to have a lower number of drug overdose deaths.

# Discussion

__Conclusions__

Based on the results, we were able to conclude that some key factors that can prevent most drug overdose deaths are:

Lower unemployment rate
High high school completion rate
Low median household income

The only factor that did not have a clear effect was some college experience.

__Limitations__

Although given the county rankings data set that was very extensive, there was no variables involving drugs or drug overdose deaths. This lead to searching for external data, which no of had county level data. This is why the data was grouped by states instead of county.

__Future Works__

Some further research that can be done is obtaining data on a county level to broaden our EDA and clusters.
Another topic that can be branched from this research is looking directly into the top five states and why they have such high drug overdose deaths. We discovered that opioids are the most used drug these cases. Looking more into what the opioid crisis looks like in each state can be an interesting project.

# Acknowledgments

Program Advisors, TA's, Optum, CMU Stats Dept.

# References

__Median Household Income:  https://worldpopulationreview.com/state-rankings/median-household-income-by-state __

__12 Month-ending Provisional Number and Percent Change of Drug Overdose Deaths: https://www.cdc.gov/nchs/nvss/vsrr/drug-overdose-data.htm __

__County Rankings: https://www.countyhealthrankings.org/explore-health-rankings/measures-data-sources/county-health-rankings-model __