---
title: "Substance Abuse"
author: "Andre Williams"
date: '2022-07-20'
output:
  html_document: default
---

# Predicators of Drug Overdose Deaths

# Introduction

Where do most drug overdose deaths occur and what are some characteristics of those areas? Some of the socio-economic factors that were looked into were the median household income, high school completion rate, some college experience rate, and unemployment rate per state. Also some information about the most used drug in a drug overdose death.

# Data

__County Health Rankings: State, High School Completion, Some College Experience, Unemployment__

```{r include=FALSE, echo=FALSE}
library(tidyverse)

d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
county_rankings <- read.csv("~/SURE-2022/county_rankings.csv")
income <- read.csv("~/SURE-2022/MedianHouseholdIncome.csv")
```

```{r include=FALSE, echo=FALSE}
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))

OD <- d3 %>%
  arrange(State) %>%
  filter(Year %in% "2019",
         Indicator %in% "Number of Drug Overdose Deaths",
         !State %in% c("DC", "YC", "US")) %>%
  select(State, Indicator, Data.Value) %>%
  group_by(State) %>%
  summarise(OD_deaths = sum(Data.Value))

colnames(OD)[1] <- 'state'
```

```{r include=FALSE, echo=FALSE}
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))

deaths <- d3 %>%
  arrange(State) %>%
  filter(Year %in% "2019",
         Indicator %in% "Number of Deaths",
         !State %in% c("DC", "YC", "US")) %>%
  select(State, Indicator, Data.Value) %>%
  group_by(State) %>%
  summarise(Total_deaths = sum(Data.Value))

colnames(deaths)[1] <- 'state'
```

```{r echo=FALSE}
county_rankings <- county_rankings %>%
  arrange(State.Abbreviation) %>%
  filter(County.FIPS.Code %in% 0,
         !State.Abbreviation %in% c("US", "DC")) %>%
  select(State.Abbreviation, Unemployment.raw.value, Some.college.raw.value, High.school.completion.raw.value) %>%
  mutate(Unemployment.raw.value = (Unemployment.raw.value * 100),
         High.school.completion.raw.value = (High.school.completion.raw.value * 100),
         Some.college.raw.value = (Some.college.raw.value * 100))

colnames(county_rankings)[1] <- 'state'
```

```{r echo=FALSE}
colnames(income)[1] <- 'state'
income <- income %>%
  mutate(state = fct_recode(state,
                            "MD" = "Maryland",
                            "NJ" = "New Jersey",
                            "HI" = "Hawaii",
                            "MA" = "Massachusetts",
                            "CT" = "Connecticut",
                            "AK" = "Alaska",
                            "NH" = "New Hampshire",
                            "CA" = "California",
                            "VA" = "Virginia",
                            "WA" = "Washington",
                            "CO" = "Colorado",
                            "UT" = "Utah",
                            "MN" = "Minnesota",
                            "NY" = "New York",
                            "DE" = "Delaware",
                            "RI" = "Rhode Island",
                            "IL" = "Illinois",
                            "ND" = "North Dakota",
                            "WY" = "Wyoming",
                            "OR" = "Oregon",
                            "VT" = "Vermont",
                            "TX" = "Texas",
                            "WI" = "Wisconsin",
                            "PA" = "Pennsylvania",
                            "NE" = "Nebraska",
                            "IA" = "Iowa",
                            "NV" = "Nevada",
                            "KS" = "Kansas",
                            "AZ" = "Arizona",
                            "GA" = "Georgia",
                            "SD" = "South Dakota",
                            "ME" = "Maine",
                            "MI" = "Michigan",
                            "OH" = "Ohio",
                            "IN" = "Indiana",
                            "ID" = "Idaho",
                            "FL" = "Florida",
                            "MO" = "Missouri",
                            "MT" = "Montana",
                            "NC" = "North Carolina",
                            "TN" = "Tennessee",
                            "SC" = "South Carolina",
                            "OK" = "Oklahoma",
                            "KY" = "Kentucky",
                            "AL" = "Alabama",
                            "NM" = "New Mexico",
                            "LA" = "Louisiana",
                            "AR" = "Arkansas",
                            "WV" = "West Virginia",
                            "MS" = "Mississippi"))
```

__Median Household Income: State, Household Income__

```{r echo=FALSE}
library(usmap)
library(ggplot2)

plot_usmap(data = income, values = "HouseholdIncome", color = "red") + 
  scale_fill_continuous(name = "Income", label = scales::comma) +
  labs(title = "Median Household Income by State") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))

```

```{r echo=FALSE, include=FALSE}
library(dplyr)

first <- dplyr::left_join(x = OD, y = deaths, by = 'state')
second <- dplyr::left_join(x = first, y = county_rankings, by = 'state')
combined <- dplyr::left_join(x = second, y = income, by = 'state')
```

__12 Month-ending Provisional Number and Percent Change of Drug Overdose Deaths: State, Indicator, Data Value__

```{r echo=FALSE}
plot_usmap(data = combined, values = "OD_deaths", color = "red") + 
  scale_fill_continuous(low = "yellow", high = "darkred", name = "Total Drug OD Deaths", label = scales::comma) +
  labs(title = "Total Drug Overdose Deaths per State") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))
```

```{r echo=FALSE, warning = FALSE, message=FALSE}
library(ggrepel)
library(wesanderson)
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))
d3[is.na(d3)] = 0

d3 <- d3 %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                        "Number of Drug Overdose Deaths"),
         State %in% "CA") %>%
  select(State, Indicator, Data.Value) %>%
  group_by(State, Indicator) %>%
  summarise(Data.Value = sum(Data.Value)) %>%
  arrange(desc(Data.Value))

df2 <- d3 %>% 
  mutate(csum = rev(cumsum(rev(Data.Value))), 
         pos = Data.Value/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Data.Value/2, pos))

d3 %>%
  ggplot(aes(x = "" , y = Data.Value, 
             fill = fct_inorder(str_wrap(Indicator, 20)))) +
  labs(title = "California's Leading Drug in Overdose Deaths") +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Paired") +
  geom_label_repel(data = df2,
                   aes(y = pos, label = paste0(Data.Value)),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Drug Type", ncol = 2, 
                             title.position = "top",
                             title.theme = element_text(size = 10),
                             label.theme = element_text(size = 7))) +
  theme_void() + 
  theme(legend.position = "right")
```

```{r echo=FALSE, warning = FALSE, message=FALSE}
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))
d3[is.na(d3)] = 0

d3 <- d3 %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                        "Number of Drug Overdose Deaths"),
         State %in% "FL") %>%
  select(State, Indicator, Data.Value) %>%
  group_by(State, Indicator) %>%
  summarise(Data.Value = sum(Data.Value)) %>%
  arrange(desc(Data.Value))

df2 <- d3 %>% 
  mutate(csum = rev(cumsum(rev(Data.Value))), 
         pos = Data.Value/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Data.Value/2, pos))

d3 %>%
  ggplot(aes(x = "" , y = Data.Value, 
             fill = fct_inorder(str_wrap(Indicator, 20)))) +
  labs(title = "Florida's Leading Drug in Overdose Deaths") +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Paired") +
  geom_label_repel(data = df2,
                   aes(y = pos, label = paste0(Data.Value)),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Indicator", ncol = 2, 
                             title.position = "top",
                             title.theme = element_text(size = 10),
                             label.theme = element_text(size = 7))) +
  theme_void() + 
  theme(legend.position = "right")
```

```{r echo=FALSE, warning = FALSE, message=FALSE}
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))
d3[is.na(d3)] = 0

d3 <- d3 %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                        "Number of Drug Overdose Deaths"),
         State %in% "TX") %>%
  select(State, Indicator, Data.Value) %>%
  group_by(State, Indicator) %>%
  summarise(Data.Value = sum(Data.Value)) %>%
  arrange(desc(Data.Value))

df2 <- d3 %>% 
  mutate(csum = rev(cumsum(rev(Data.Value))), 
         pos = Data.Value/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Data.Value/2, pos))

d3 %>%
  ggplot(aes(x = "" , y = Data.Value, 
             fill = fct_inorder(str_wrap(Indicator, 20)))) +
  labs(title = "Texas' Leading Drug in Overdose Deaths") +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Paired") +
  geom_label_repel(data = df2,
                   aes(y = pos, label = paste0(Data.Value)),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Indicator", ncol = 2, 
                             title.position = "top",
                             title.theme = element_text(size = 10),
                             label.theme = element_text(size = 7))) +
  theme_void() + 
  theme(legend.position = "right")
```

```{r echo=FALSE, warning = FALSE, message=FALSE}
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))
d3[is.na(d3)] = 0

d3 <- d3 %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                        "Number of Drug Overdose Deaths"),
         State %in% "OH") %>%
  select(State, Indicator, Data.Value) %>%
  group_by(State, Indicator) %>%
  summarise(Data.Value = sum(Data.Value)) %>%
  arrange(desc(Data.Value))
df2 <- d3 %>% 
  mutate(csum = rev(cumsum(rev(Data.Value))), 
         pos = Data.Value/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Data.Value/2, pos))

d3 %>%
  ggplot(aes(x = "" , y = Data.Value, 
             fill = fct_inorder(str_wrap(Indicator, 20)))) +
  labs(title = "Ohio's Leading Drug in Overdose Deaths") +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Paired") +
  geom_label_repel(data = df2,
                   aes(y = pos, label = paste0(Data.Value)),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Indicator", ncol = 2, 
                             title.position = "top",
                             title.theme = element_text(size = 10),
                             label.theme = element_text(size = 7))) +
  theme_void() + 
  theme(legend.position = "right")
```

# Methods

__K-Means Clustering__

I implemented K-means clustering to determine if there were any correlation between a drug overdose death and the socio-economic factors studied.

```{r echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
init_kmeans <- 
  kmeans(dplyr::select(combined,
                       OD_deaths, Unemployment.raw.value),
         algorithm = "Lloyd", centers = 3,
         nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(init_kmeans$cluster)) %>%
  ggplot(aes(x = OD_deaths, y = Unemployment.raw.value,
             color = state_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r echo=FALSE, include=FALSE,warning=FALSE, message=FALSE}
init_kmeans <- 
  kmeans(dplyr::select(combined,
                       OD_deaths, Some.college.raw.value),
         algorithm = "Lloyd", centers = 3,
         nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(init_kmeans$cluster)) %>%
  ggplot(aes(x = OD_deaths, y = Some.college.raw.value,
             color = state_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r echo=FALSE, include=FALSE,warning=FALSE, message=FALSE}
init_kmeans <- 
  kmeans(dplyr::select(combined,
                       OD_deaths, High.school.completion.raw.value),
         algorithm = "Lloyd", centers = 3,
         nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(init_kmeans$cluster)) %>%
  ggplot(aes(x = OD_deaths, y = High.school.completion.raw.value,
             color = state_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Results

```{r echo=FALSE}
library(ggrepel)
combined <- combined %>%
  mutate(std_death = as.numeric(scale(OD_deaths, center = TRUE, scale = TRUE)),
         std_unemploy = as.numeric(scale(Unemployment.raw.value, center = TRUE, scale = TRUE)),
         std_hs = as.numeric(scale(High.school.completion.raw.value, center = TRUE, scale = TRUE)),
         std_college = as.numeric(scale(Some.college.raw.value, center = TRUE, scale = TRUE)))
```

```{r echo=FALSE,warning=FALSE, message=FALSE}
std_kmeans <- 
  kmeans(dplyr::select(combined, std_death, std_unemploy),
         algorithm = "Lloyd", centers = 4, nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(std_kmeans$cluster)) %>%
  ggplot(aes(x = std_death, y = std_unemploy,
             color = state_clusters, label = state)) +
  labs(x = "Total Deaths", y = "Unemployment Rate",
       title = "Cluster of Deaths and Unemployment Rate by State") +
  geom_point() +
  geom_text_repel(size = 2) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5)) +
  coord_fixed()

```

```{r echo=FALSE,warning=FALSE, message=FALSE}
std_kmeans <- 
  kmeans(dplyr::select(combined, std_death, std_hs),
         algorithm = "Lloyd", centers = 4, nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(std_kmeans$cluster)) %>%
  ggplot(aes(x = std_death, y = std_hs,
             color = state_clusters, label = state)) +
  labs(x = "Total Deaths", y = "HS Completion",
       title = "Cluster of Deaths and HS Completion Rate by State") +
  geom_point() + 
  geom_text_repel(size = 2) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5)) +
  coord_fixed()

```

```{r echo=FALSE,warning=FALSE, message=FALSE}
std_kmeans <- 
  kmeans(dplyr::select(combined, std_death, std_college),
         algorithm = "Lloyd", centers = 4, nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(std_kmeans$cluster)) %>%
  ggplot(aes(x = std_death, y = std_college,
             color = state_clusters, label = state)) +
  labs(x = "Total Deaths", y = "College Experience",
       title = "Cluster of Deaths and College Experience Rate by State") +
  geom_point() + 
  geom_text_repel(size = 2) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5)) +
  coord_fixed()

```