---
title: "Substance Abuse"
author: "Andre Williams"
date: '2022-07-20'
output: html_document
---

# Hypotheses 1: Understanding where most drug overdose deaths occur and some characteristics of those areas


```{r include=FALSE, echo=FALSE}
library(tidyverse)

d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
county_rankings <- read.csv("~/SURE-2022/county_rankings.csv")
```

```{r include=FALSE, echo=FALSE}
# Cleaning the data

d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))

d3 <- d3 %>%
  arrange(State) %>%
  filter(Year %in% "2019",
         Indicator %in% "Number of Drug Overdose Deaths",
         !State %in% c("DC", "YC", "US")) %>%
  select(State, Indicator, Data.Value) %>%
  group_by(State) %>%
  summarise(total_deaths = sum(Data.Value))

colnames(d3)[1] <- 'state'
```

```{r include=FALSE, echo=FALSE}
county_rankings <- read.csv("~/SURE-2022/county_rankings.csv")

county_rankings <- county_rankings %>%
  arrange(State.Abbreviation) %>%
  filter(County.FIPS.Code %in% 0,
         !State.Abbreviation %in% c("US", "DC")) %>%
  select(State.Abbreviation, Unemployment.raw.value, Some.college.raw.value, High.school.completion.raw.value)

colnames(county_rankings)[1] <- 'state'
```

```{r}
# Combining the data

library(dplyr)

combined <- dplyr::left_join(x = d3, y = county_rankings, by = 'state')
```

__Plotting the total drug overdose deaths per state in 2019__

```{r echo=FALSE}
library(usmap)
library(ggplot2)

plot_usmap(data = combined, values = "total_deaths", color = "red") + 
  scale_fill_continuous(low = "yellow", high = "darkred", name = "Total Deaths", label = scales::comma) + 
  theme(legend.position = "right")

```

__Looking at the most used drug in California__

```{r echo=FALSE}
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))

d3 <- d3 %>%
  arrange(desc(Data.Value)) %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                        "Number of Drug Overdose Deaths"),
         State %in% "CA") %>%
  select(State, Indicator, Data.Value)

head(d3, 3)
```

__Looking at the most used drug in Florida__

```{r echo=FALSE}
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))

d3 <- d3 %>%
  arrange(desc(Data.Value)) %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                        "Number of Drug Overdose Deaths"),
         State %in% "FL") %>%
  select(State, Indicator, Data.Value)

head(d3, 3)
```

__Looking at the most used drug in Texas__

```{r echo=FALSE}
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))

d3 <- d3 %>%
  arrange(desc(Data.Value)) %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                        "Number of Drug Overdose Deaths"),
         State %in% "TX") %>%
  select(State, Indicator, Data.Value)

head(d3, 3)
```

__Looking at the most used drug in Ohio__

```{r echo=FALSE}
d3 <- read.csv("~/SURE-2022/Drug-Alcohol Overdose Deaths.csv")
d3$Data.Value <- as.integer(gsub(",", "", d3$Data.Value))

d3 <- d3 %>%
  arrange(desc(Data.Value)) %>%
  filter(Year %in% "2019",
         !Indicator %in% c("Percent with drugs specified", "Number of Deaths",
                           "Number of Drug Overdose Deaths"),
         State %in% "OH") %>%
  select(State, Indicator, Data.Value)

head(d3, 3)
```

__Median Household Income By State__

```{r echo=FALSE}
income <- read.csv("~/SURE-2022/MedianHouseholdIncome.csv")

library(usmap)
library(ggplot2)


colnames(income)[1] <- 'state'

plot_usmap(data = income, values = "HouseholdIncome", color = "red") + 
  scale_fill_continuous(low = "yellow", high = "darkred", name = "Median Income", label = scales::comma) + 
  theme(legend.position = "right")

```


__Highschool Completion__

```{r echo=FALSE}
library(usmap)
library(ggplot2)

plot_usmap(data = combined, values = "High.school.completion.raw.value",
           color = "red") + 
  scale_fill_continuous(low = "yellow", high = "darkred", name = "HS Completion %", label = scales::comma) + 
  theme(legend.position = "right")
```

__Unemployment Rate by State__

```{r echo=FALSE}
library(usmap)
library(ggplot2)

plot_usmap(data = combined, values = "Unemployment.raw.value",
           color = "red") + 
  scale_fill_continuous(low = "yellow", high = "darkred", name = "Unemployment %", label = scales::comma) + 
  theme(legend.position = "right")
```

__College Experience__

```{r echo=FALSE}
library(usmap)
library(ggplot2)

plot_usmap(data = combined, values = "Some.college.raw.value",
           color = "red") + 
  scale_fill_continuous(low = "yellow", high = "darkred", name = "College Experience %", label = scales::comma) + 
  theme(legend.position = "right")
```

_Clustering of Total deaths by the socioeconomic factors__

```{r echo=FALSE}
init_kmeans <- 
  kmeans(dplyr::select(combined,
                       total_deaths, Unemployment.raw.value),
         algorithm = "Lloyd", centers = 3,
         nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(init_kmeans$cluster)) %>%
  ggplot(aes(x = total_deaths, y = Unemployment.raw.value,
             color = state_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r echo=FALSE}
init_kmeans <- 
  kmeans(dplyr::select(combined,
                       total_deaths, Some.college.raw.value),
         algorithm = "Lloyd", centers = 3,
         nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(init_kmeans$cluster)) %>%
  ggplot(aes(x = total_deaths, y = Some.college.raw.value,
             color = state_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r echo=FALSE}
init_kmeans <- 
  kmeans(dplyr::select(combined,
                       total_deaths, High.school.completion.raw.value),
         algorithm = "Lloyd", centers = 3,
         nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(init_kmeans$cluster)) %>%
  ggplot(aes(x = total_deaths, y = High.school.completion.raw.value,
             color = state_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")
```

__Clustering with the standardized variables__

```{r echo=FALSE}
combined <- combined %>%
  mutate(std_death = as.numeric(scale(total_deaths, center = TRUE, scale = TRUE)),
         std_unemploy = as.numeric(scale(Unemployment.raw.value, center = TRUE, scale = TRUE)),
         std_hs = as.numeric(scale(High.school.completion.raw.value, center = TRUE, scale = TRUE)),
         std_college = as.numeric(scale(Some.college.raw.value, center = TRUE, scale = TRUE)))
```

```{r echo=FALSE}
std_kmeans <- 
  kmeans(dplyr::select(combined, std_death, std_unemploy),
         algorithm = "Lloyd", centers = 4, nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(std_kmeans$cluster)) %>%
  ggplot(aes(x = std_death, y = std_unemploy,
             color = state_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_fixed()

```

```{r echo=FALSE}
std_kmeans <- 
  kmeans(dplyr::select(combined, std_death, std_hs),
         algorithm = "Lloyd", centers = 4, nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(std_kmeans$cluster)) %>%
  ggplot(aes(x = std_death, y = std_hs,
             color = state_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_fixed()

```

```{r echo=FALSE}
std_kmeans <- 
  kmeans(dplyr::select(combined, std_death, std_college),
         algorithm = "Lloyd", centers = 4, nstart = 1)

combined %>%
  mutate(state_clusters = 
           as.factor(std_kmeans$cluster)) %>%
  ggplot(aes(x = std_death, y = std_college,
             color = state_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_fixed()

```

__Model Based Clustering__

```{r echo=FALSE, include=FALSE}
library(mclust)

death_college_clust <- Mclust(dplyr::select(combined, total_deaths, Some.college.raw.value))

death_unemploy_clust <- Mclust(dplyr::select(combined, total_deaths, Unemployment.raw.value))

death_hs_clust <- Mclust(dplyr::select(combined, total_deaths, High.school.completion.raw.value))
```

```{r echo=FALSE}
plot(death_college_clust, what = 'classification')
```

```{r echo=FALSE}
plot(death_hs_clust, what = 'classification')
```

```{r echo=FALSE}
plot(death_unemploy_clust, what = 'classification')
```
